\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-03-06 analysis/paper/paper.Rmd
NOT COMPLETE
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags
add lines to dot whisker plot
Need to removeRcolorBrewer palletes

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
#devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
library(pracma)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
```

@@

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}

completeData<-casebase::support
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])

@@

Our next step is to get the hazard function. All covariates will be used to estimate the hazard function. The hazard function will be used to estimate the cumulative incidence of each individual in the study. Then, the average cumulative incidience curve for all observations will be used for our comparison.

Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
#devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
library(pracma)
library(RColorBrewer)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- FALSE
eval_theory <- FALSE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
support<-casebase::support

@@

Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a simple Kaplan-meier curve. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. We will then compare each modeling using the IPA score (???NEED REFERENCE HERE???) We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.  Over 5.56 years, \ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA??? has the best calibration and discrimination with an IPA score of ???INSERT ACTUAL BEST SCORE??? . ??BEST MODEL?? will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.

```{r baselinePlotsPrep, eval=eval_cs3}
completeData<-casebase::support
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])

fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),x))
train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
test_index <- setdiff(1:nrow(fullDataCox), train_index)

train<-fullDataCox[train_index,]
test<-fullDataCox[test_index,]
coxSPS<- survival::coxph(Surv(time=V1,event=V2) ~ sps, data = train)
abcoxSPS<-survival::survfit(coxSPS,newdata = test)

coxAPS<- survival::coxph(Surv(time=V1,event=V2) ~ aps, data = train)
abcoxAPS<-survival::survfit(coxAPS,newdata = test)

coxFull<- survival::coxph(Surv(time=V1,event=V2) ~ .-sps -aps, data = train)
abcoxFull<-survival::survfit(coxAPS,newdata = test)
#library(prodlim)

KM<- survival::coxph(Surv(time=V1,event=V2) ~ 1, data = test)
abKM<-survival::survfit(KM)
#abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
abKMcompare<-cbind(abKM$time[abKM$time<=1631],1-abKM$surv[abKM$time<=1631])
colnames(abKMcompare)<-c("Time","KM")

baseCoxComparisonData<-cbind(abcoxSPS$time[abcoxSPS$time<=1631],1-rowMeans(abcoxSPS$surv[abcoxSPS$time<=1631,]),1-rowMeans(abcoxAPS$surv[abcoxSPS$time<=1631,]),1-abKM$surv,1-rowMeans(abcoxFull$surv[abcoxSPS$time<=1631,]))
colnames(baseCoxComparisonData)<-c("Time","SPS","APS","Unadjusted","Full")
myColors <- brewer.pal(5,"Dark2")
FigureWidth='90%'

```

```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} AbsoluteRisk curves from hazard models. Pink: All covariates without SPS nor APS. Green: APS only model. Blue: SPS only model. Purple: unadjusted curve. All three models underestimate the risk in comparison to the unadjusted curve."}
ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) + 
  geom_line(aes(y = SPS, colour = "SPS"),lwd=1) + 
  geom_line(aes(y = APS, colour = "APS"),lwd=1) +
  geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=1)+
  geom_line(aes(y = Full, col = "All Covariates without SPS APS"),lwd=1)+
  labs(y= "Probability of Death", x = "Survival time (Days)")
```

To get the absolute risk following regularized cox regression, this required that we work around limitations in existing packages. To make use of the \pkg{survival} package, we first get our hazard coefficients using \pkg{glmnet}'s coxnet implementation for lasso penalization [@regpathcox]. Next, we fit a cox model using the survival package with the selected variables only, from our previous fit. Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.

For our penalized logistic regression model, we opted for a logistic function of time, resulting in a Weibull distribution assumption. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function, depicted in the ??? curve in \ref{fig:cs3Final}.
??? I AM STILL PENALIZING TIME ???

We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time. These individuals are then average resulting in the ???curve in \ref{fig:cs3Final}. ???SAME BUG IS BACK???

```{r supportSetUp, echo = FALSE, eval = eval_cs3}

ratio=10  #universal ratio for TRUE
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), )
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]

#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(train[,c(2,1)])
colnames(y)<-c("death","d.time")
x=data.matrix(train[,-c(2,1)])
newData=data.matrix(test[,-c(2,1)])
```

```{r coxHazAbsolute, echo = FALSE,eval= eval_cs3 }
#Create u and xCox, which will be used when fitting Cox with glmnet.

u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))

xCox=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=length(x[,1]))),x))
colnames(xCox)[2]="(Intercept)"

#hazard for cox using glmnet
coxNet=glmnet::cv.glmnet(x=x,y=u, family="cox",alpha=1,standardize = TRUE)
#convergence demonstrated in plot
#plot(coxNet)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxData<-as.data.frame(cbind(d.time=as.numeric(y[,2]),death=as.numeric(y[,1]),x[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxNet <- survival::coxph(Surv(time=d.time,event=death) ~ ., data = cleanCoxData)

#extract coefficients for lollipop plot later
estimatescoxNet=setDT(as.data.frame(coef(coxNet)), keep.rownames = TRUE)[]
#estimatescoxNet$std.error<-0
estimatescoxNet$Model="CoxNet"
colnames(estimatescoxNet)<-c("term","estimate","model")

#The coefficients of this object will be replaced with the estimates from coxNet.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxNet$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
#posCox=nonzero_covariate_cox$X1
#newDataCoxI=cleanCoxDataI[,posCox]

abcoxNet<-survival::survfit(coxNet,type="breslow",newdata=as.data.frame(newData[,nonzero_covariate_cox$X1]))

abCN=as.data.frame(cbind(abcoxNet$time, 1-rowMeans(abcoxNet$surv)))
colnames(abCN)<-c("Time","Absolute risk")
abCN$Model="Coxnet"
abCN=abCN[abCN$Time<=max(abKMcompare[,1]),]
#plot(abCN[,1],1-abCN[,2])
```

```{r cbHazard}, echo = FALSE, eval = eval_cs3}

cbFull=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",formula_time = ~log(d.time),alpha=1,ratio=ratio,standardize = TRUE)
```

```{r CBabsolute, echo = FALSE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
abCbFull=casebase::absoluteRisk(cbFull,time = seq(1,max(abKMcompare[,1]), 25),newdata = newData , s="lambda.1se",method=c("numerical"))
abCB=data.frame(abCbFull[,1],rowMeans(abCbFull[,-c(1)]))

colnames(abCB)<-c("Time","Absolute risk")
abCB$Model="Penalized logistic"

ab=rbind(abCB,abCN)

```

```{r include=TRUE,eval=eval_cs3,fig.cap="\\label{fig:baselines} HERES SOME TEXT ABT IT"}
ggplot(as.data.frame(ab), aes(Time)) + 
    geom_line(data=as.data.frame(baseCoxComparisonData),mapping=aes(y = Full, colour = "Cox"),lwd=2)+
  geom_line(aes(y = `Absolute risk`, colour = Model),lwd=2)+
  geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=2)+
  scale_colour_manual(name = "Models",values = myColors[c(4,1,3,2)])+
  labs(y= "Probability of Death", x = "Survival time (Days)")
```

???SHORT SPIEL ABOUT IPA BETWEEN SEMI_PARAMETRIC AND PARAMETRIC which leads to conclusion of best overall model. This refers to the COEFPLOT in a following figure???

```{r coefplots, eval=eval_cs3}

library(dotwhisker)
library(broom)
library(dplyr)

estimatescb=setDT(as.data.frame(coef(cbFull)[-c(1,2),1]), keep.rownames = TRUE)[]
#estimatescb$std.error<-0
estimatescb$Model="casebase"
colnames(estimatescb)<-c("term","estimate","model")

estimatescoxFull=setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
#estimatescoxFull$std.error<-0
estimatescoxFull$Model="CoxFull"
colnames(estimatescoxFull)<-c("term","estimate","model")

lolliplotDots<-rbind(estimatescb,estimatescoxNet,estimatescoxFull)
lolliplotDots$conf.low<-NA
lolliplotDots$conf.high<-NA
lolliplotDots$conf.high[lolliplotDots$estimate< 0]<-0

lolliplotDots$conf.low[lolliplotDots$estimate> 0]<-0

dwplot(lolliplotDots)

```
